---
title: "Methodological Background: Confidence Limits for Prediction Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodological Background: Confidence Limits for Prediction Performance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette provides the methodological background for `pppms`. It is
written as a bridge between the package interface and the statistical ideas
developed in the dissertation *Confidence Limits for Prediction Performance*.

The package addresses the following inferential problem: given several
candidate prediction rules, and after selecting the empirically best one on
the basis of the same evaluation data, how can one quantify uncertainty for
the prediction performance of the selected rule while accounting for the
selection step?

The method implemented here is based on **multiplicity-adjusted bootstrap
tilting** and produces **lower confidence limits** for the prediction
performance of the selected rule.

# Statistical setup

Suppose an evaluation sample of size \(n\) is available, with binary outcomes

\[
Y_1, \dots, Y_n \in \{0,1\},
\]

and \(K\) candidate prediction rules. For candidate \(j \in \{1,\dots,K\}\),
let \(\hat Y_{ij}\) denote the prediction for observation \(i\).

For classification accuracy, define the observation-wise contribution

\[
Z_{ij} = I(Y_i = \hat Y_{ij}),
\]

so that the empirical performance of candidate \(j\) is

\[
\hat \theta_j = \frac{1}{n}\sum_{i=1}^n Z_{ij}.
\]

The empirically best candidate is then

\[
\hat j = \arg\max_{1 \le j \le K} \hat \theta_j.
\]

A naive analysis would report \(\hat \theta_{\hat j}\), possibly together
with a standard confidence interval for a fixed model. This is generally not
valid after model selection, because the same data have been used both to
identify the best candidate and to evaluate it.

# Why the naive analysis is problematic

Selection among several candidates introduces optimism. Even when each
candidate is assessed fairly on its own, the maximum over many noisy estimates
tends to be too large. This is a multiplicity phenomenon.

The key inferential challenge is therefore not merely to estimate a single
performance parameter, but to perform valid inference **after the best model
has been selected from a finite set of competitors**.

This naturally leads to a simultaneous-inference perspective: one should
account for the whole family of candidate models rather than treating the
selected one as if it had been chosen in advance.

# Conditional prediction performance

The target of inference is the performance of the selected rule, conditional
on the concrete set of candidate prediction rules under consideration.

This is the practically relevant target in many applied settings. Analysts do
not deploy a random model drawn from a model-generating mechanism. They deploy
the rule that emerges from a specific comparison or selection process. The
selection step is therefore part of the inferential problem.

The goal is to construct a lower confidence bound \(L_\alpha\) such that,
approximately,

\[
\Pr\!\left(\theta_{\hat j} \ge L_\alpha\right) \ge 1 - \alpha,
\]

where \(\theta_{\hat j}\) denotes the prediction performance of the selected
candidate rule.

# Multiplicity-adjusted bootstrap tilting

The package combines two ideas:

1. **multiplicity adjustment**, to account for simultaneous comparison across
   candidate models;
2. **bootstrap tilting**, to evaluate the tail behavior relevant for a lower
   confidence bound.

## Multiplicity adjustment

A max-type calibration is used to account for the fact that several candidate
models are compared. Rather than calibrating uncertainty model by model, the
procedure works with the joint bootstrap distribution of standardized
statistics across all candidates.

This prevents the selected rule from being treated as if it had been fixed in
advance.

## Bootstrap tilting

In ordinary bootstrap resampling, observations receive equal resampling
probability. In tilted bootstrap procedures, these probabilities are modified
through weights of the form

\[
w_i(\tau) \propto \exp\!\bigl(\tau \psi_i\bigr),
\]

where \(\psi_i\) is an empirical influence quantity and \(\tau\) is a tilting
parameter.

The purpose of tilting is to move the resampling distribution toward the
boundary case relevant for the lower confidence limit. This can improve the
stability and relevance of the calibration for tail probabilities.

In implementation, the weights are usually normalized after working on the log
scale for numerical stability.

# High-level algorithm

The function `MabtCI()` implements the following logic.

1. Compute empirical prediction performance for each candidate model.
2. Select the empirically best model.
3. Generate stratified bootstrap samples from the evaluation data.
4. Recompute model-wise performance estimates in each bootstrap sample.
5. Standardize the bootstrap deviations to obtain a joint bootstrap sample of
   statistics.
6. Transform the bootstrap sample into a multiplicity-adjusted reference
   distribution based on maxima across candidate models.
7. Introduce tilted observation weights indexed by the tilting parameter
   \(\tau\).
8. Evaluate a p-value-like calibration function under the tilted bootstrap
   distribution.
9. Solve for the value of \(\tau\) such that the calibration matches the
   nominal significance level \(\alpha\).
10. Convert the corresponding tilted mean into the lower confidence bound.

# Relation to the package interface

The current package interface is intentionally narrow:

```{r eval = FALSE}
res <- MabtCI(
  true_labels = y,
  pred_labels = pred_mat,
  alpha = 0.05,
  B = 10000,
  seed = 1
)
```

The package expects:

- `true_labels`: a binary outcome vector coded as `0` and `1`;
- `pred_labels`: a matrix whose columns represent candidate prediction rules;
- `alpha`: the target significance level;
- `B`: the number of bootstrap replications;
- `seed`: an optional random seed.

The returned object currently contains:

- `bound`: the lower confidence bound;
- `tau`: the estimated tilting parameter;
- `t0`: the observed performance of the selected candidate;
- `selected_idx`: the index of the selected candidate.

This interface emphasizes the inferential end product rather than exposing all
internal bootstrap objects. That is deliberate: `pppms` is meant to be a
methods package rather than a general prediction framework.

# Example

The following toy example illustrates the package workflow.

```{r example}
library(pppms)

y <- c(0, 0, 1, 1, 0, 1)

pred_mat <- cbind(
  model1 = c(0, 0, 1, 1, 1, 1),
  model2 = c(0, 1, 1, 0, 0, 1)
)

res <- MabtCI(
  true_labels = y,
  pred_labels = pred_mat,
  alpha = 0.05,
  B = 200,
  seed = 1
)

res
```

A typical interpretation is:

- `t0` is the observed empirical performance of the selected model;
- `tau` is the tilting parameter solving the calibration equation;
- `bound` is the multiplicity-adjusted lower confidence limit for the selected
  model's prediction performance.

# Interpretation of the lower bound

The lower confidence bound should not be read as a corrected point estimate.
Its role is inferential.

For example, if the method returns a lower bound of 0.78 at confidence level
\(1-\alpha = 0.95\), the intended interpretation is that the data support the
claim that the prediction performance of the selected rule is at least 0.78,
up to the approximation error inherent in the method.

This is especially useful in settings where several candidate models perform
similarly and naive performance reporting would exaggerate the evidence in
favor of the empirically best one.

# Scope and limitations

The current implementation is deliberately focused. It provides a working and
interpretable first version of the core method for binary prediction settings
with candidate prediction rules represented by columns of a prediction matrix.

This package should make clear which parts are already implemented and which
extensions would require additional methodological work. In particular,
generalization to more complex performance measures or dependence structures
may require theory beyond the currently implemented version.

# Positioning

`pppms` is not intended as a machine-learning framework for model fitting.
Its contribution is inferential.

The package is meant for users who already have candidate prediction rules and
want principled uncertainty quantification for the selected rule's performance,
with explicit adjustment for the model-selection step.

# References

Rink, P. (2025). *Confidence Limits for Prediction Performance*. Doctoral
thesis, University of Bremen.
